GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2025-06-12 writeUps/paper.tex

\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
%\RequirePackage[authoryear]{natbib}
\usepackage{longtable}
%\usepackage[notablist,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

@@

Data from the randomized controlled trial (RCT) reported in \citet{impactPaper} may provide evidence for one or another side of this debate.
That RCT contrasted four different applications for computer-based math practice, one of which included bottom-out hints.
If that condition benefited students who frequently requested bottom-out hints (``bottom-outers'') more than those who didn't, that may suggest that the availability of bottom-out hints played a role in the condition's effectiveness.
Alternatively, if students who avoided bottom-out hints benefited more, it may be because the availability of bottom-out hints impeded the condition's effectiveness.
(There are certainly other factors at play as well, so this analysis would not be conclusive.)

Estimating an effect for bottom-outers is a tricky proposition.
Causal inference in statistics typically depends on comparisons of outcomes between similar groups of subjects. However, there is no readily identifiable comparison group for bottom-outers, who are only represented in one of the four treatment conditions, and who presumably struggled more frequently, on average, than their non-bottom-outer peers.
We will formulate a solution using Principal stratification \citep{frangakis}, a framework for defining and estimating effects for post-treatment subgroups, that is, subgroups that may themselves be a function of treatment assignment. It relies on the notion of \emph{potential} subgroups called principal strata: which subgroup would a subject be in if they were (perhaps counterfactually) assigned to one treatment condition or another.
For example, the appropriate comparison group for bottom-outers is the set of students, assigned to the other treatment arms, who \emph{would have} requested bottom-out hints if only given that option.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=5.5in,clip]{../simFigs/boxplots.pdf}
  \caption{Violin and scatter plots of 500 simulation estimation errors for estimators described in Section~\ref{sec:simMods} under varying conditions. For all plots, $n=500$; there were 500 replications for each set of conditions. Annotations indicate the number of extreme outliers excluded from the plotting area.}
  \label{fig:boxplots}
\end{figure}

Figure \ref{fig:boxplots} shows violin plots, layered on top of jittered scatter plots, showing estimation error for \textsc{geepers}, the mixture estimator, and the \textsc{psw} estimator of $\eff1$ under various sets of conditions, and Table \ref{tab:coverage} shows empirical coverage of nominal 95\% confidence interval estimates under a somewhat wider set of circumstances when $\muc1=0.3$. More complete results, including root-mean-squared-error estimates, are in the online appendix.
The leftmost panel of Figure \ref{fig:boxplots} and the top line of Table \ref{tab:coverage} show results under conditions favorable to both the mixture model and to \textsc{geepers}---normal residuals and no interactions in the data generating model between covariates and either $S$ or $Z$.
\textsc{geepers} was roughly unbiased, but less precise than the other two methods---when $\alpha=0.2$ it was much noisier, and when $\alpha=0.5$ it was only slightly less precise.
The mixture estimator's bias depended on $\muc1$---when $\muc1=0$, the bias was slight, and when $\muc1=0.3$, the bias was more substantial. This result is somewhat surprising, because when $\muc1=0$, there was no separation between the principal strata in the control group, and previous literature \citep{griffin2008application} has identified this scenario as a particular challenge for mixture modeling.
The \textsc{psw} estimator was also biased, due, presumably, to the violation of Principal Ignorability, Assumption \ref{ass:PI}, inherent in the data generating model. On the other hand, the \textsc{psw} estimator had easily the lowest sampling variance of the three estimators.

The middle panel of Figure \ref{fig:boxplots} and the fifth line of Table \ref{tab:coverage} show results for when the residuals in \eqref{eq:y-sim} are uniformly distributed.
In this case, \textsc{geepers} and \textsc{psw} estimators, which do not assume normality, behaved in a similar fashion as in the normal case, but the mixture estimator showed a bimodal pattern--sometimes over-estimating $\eff1$ and sometimes underestimating, but rarely estimating $\eff1$ accurately.
Credible intervals from the mixture model exhibited severe undercoverage, while \textsc{geepers} confidence intervals achieved at least their nominal level.

%\sloppy
The rightmost panel of Figure \ref{fig:boxplots} returns to the case of normal residuals---and only shows results when $\muc1=0.3$ and $\alpha=0.5$---but introduces interactions in the data generating model between covariates and either $\st$, $Z$, or both, violating Assumption \ref{ass:rci}.
%Presence of interactions could undermine the \textsc{geepers} by inducing a relationship between principal scores and potential outcomes, even after adjusting the outcomes for covariate effects.
Fortunately, these interactions made little difference when $\alpha=0.5$. However, Table \ref{tab:coverage} shows that when $\alpha=0$ or 0.2, and interactions between $\bx$ and $Z$ were present (i.e., $\gamma_3\ne 0$), \textsc{geepers} 95\% confidence intervals tended to under-cover. Nevertheless, coverage of \textsc{geepers} intervals was nearly always substantially higher than coverage of corresponding 95\% credible intervals from mixture models.%\footnote{}

@@

>>>>

\RequirePackage{amsmath,amsfonts,amssymb}
%\RequirePackage[authoryear]{natbib}
\usepackage{longtable}
%\usepackage[notablist,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

@@

Data from the randomized controlled trial (RCT) reported in \citet{impactPaper} may provide evidence for one or another side of this debate.
That RCT contrasted four different applications for computer-based math practice, one of which included bottom-out hints.
If that condition benefited students who frequently requested bottom-out hints (``bottom-outers'') more than those who didn't, that may suggest that the availability of bottom-out hints played a role in the condition's effectiveness.
Alternatively, if students who avoided bottom-out hints benefited more, it may be because the availability of bottom-out hints impeded the condition's effectiveness.
(There are certainly other factors at play as well, so this analysis would not be conclusive.)
Indeed, if the use of bottom-out hints is considered part of the treatment, then those students who rarely used them are partial compliers---could the treatment still benefit them?

Estimating an effect for bottom-outers, or other partial compliers, is a tricky proposition.
Causal inference in statistics typically depends on comparisons of outcomes between similar groups of subjects. However, there is no readily identifiable comparison group for bottom-outers, who are only represented in one of the four treatment conditions, and who presumably struggled more frequently, on average, than their non-bottom-outer peers.
We will formulate a solution using Principal stratification \citep{frangakis}, a framework for defining and estimating effects for post-treatment subgroups, that is, subgroups that may themselves be a function of treatment assignment. It relies on the notion of \emph{potential} subgroups called principal strata: which subgroup would a subject be in if they were (perhaps counterfactually) assigned to one treatment condition or another.
For example, the appropriate comparison group for bottom-outers is the set of students, assigned to the other treatment arms, who \emph{would have} requested bottom-out hints if only given that option.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=5.5in,clip]{../simFigs/boxplotsNew.pdf}
  \caption{Violin and scatter plots of 500 simulation estimation errors for estimators described in Section~\ref{sec:simMods} under varying conditions. For all plots, $n=500$; there were 500 replications for each set of conditions. Annotations indicate the number of extreme outliers excluded from the plotting area.}
  \label{fig:boxplots}
\end{figure}

%<<<<<<< HEAD
Figure \ref{fig:boxplots} shows violin plots, layered on top of jittered scatter plots, showing estimation error for \textsc{geepers}, the parametric mixture model, and the \textsc{psw} estimator of $\eff1$ for, and Table \ref{tab:coverage} shows empirical coverage of nominal 95\% confidence interval estimates for \textsc{geepers} and a parametric mixture models; when at least one of an estimator's assumptions is violated, the result is colored red.

The first row of Table \ref{tab:coverage} and the first column of Figure \ref{fig:boxplots} show similar results as Figure \ref{fig:alphan}---when all assumptions of both methods are met, \textsc{geepers} is slightly less biased than the parametric mixture model; when $\alpha=0.2$, \textsc{geepers} is much more variable and $\alpha=0.5$ the difference in variance is slight.
\textsc{psw} has by far the lowest variance and the hightest bias; the latter is due to the violation of principal ignorability.

The remainder of the results reflect violations modeling assumptions for either \geepers or the parametric mixture model or both: interactions in the data generating model between covariates $X$ and treatment assignment $Z$ and/or principal stratum $S$, either of which would violate assumption \ref{ass:rci} and the linear additive model of the \textsc{pmm}, and/or residuals drawn from a uniform distribution, which violates the normality assumption of \textsc{pmm}.
For \geepers, when $\alpha=0.2$ the presence of interactions in the data generating model increases sampling variance, but only interactions between $X$ and $Z$ reduce confidence interval coverage. When $\alpha=0.5$, the presence of interactions has little to no effect on the performance of the \geepers estimator, and all 95\% confidence intervals approximately acheive their nominal levels. \geepers makes no assumptions about the shape of the residual distribution and behaves similarly under either normal or uniform residuals.

Like \geepers, \pmm is insensitive to interactions between $X$ and $S$. However, interactions between $X$ and $Z$ in the data generating model skew the \pmm sampling distribution and reduce the coverage of its 95\% confidence intervals, even when $\alpha=0.5$.
When the residual distribution was uniformly distributed, the \pmm sampling distribution appeared bimodal, and its credible intervals exhibited severe undercoverage.

In sum, \psw is the least variable of all three estimator, and does not depend on outcome modeling, but is vulnerable to bias from an unmeasured covariate that predicts both $S$ and $Y$.
When all assumptions of both methods are met, \geepers is less efficient, but also less biased, than \pmm, but differences in efficiency decrease with $n$ and $\alpha$.
Neither \psw nor \pmm are vulnerable to interactions beetween $X$ and $S$.
\pmm is highly vulnerable to violations of normality and to interactions between $X$ and $Z$, while \psw is robust to non-normality and robust to $X$-$Z$ interactions when $\alpha-0.5$.
%=======
% Figure \ref{fig:boxplots} shows violin plots, layered on top of jittered scatter plots, showing estimation error for \textsc{geepers}, the mixture estimator, and the \textsc{psw} estimator of $\eff1$ under various sets of conditions, and Table \ref{tab:coverage} shows empirical coverage of nominal 95\% confidence interval estimates under a somewhat wider set of circumstances when $\muc1=0.3$. More complete results, including root-mean-squared-error estimates, are in the online appendix.
% The leftmost panel of Figure \ref{fig:boxplots} and the top line of Table \ref{tab:coverage} show results under conditions favorable to both the mixture model and to \textsc{geepers}---normal residuals and no interactions in the data generating model between covariates and either $S$ or $Z$.
% \textsc{geepers} was roughly unbiased, but less precise than the other two methods---when $\alpha=0.2$ it was much noisier, and when $\alpha=0.5$ it was only slightly less precise.
% The mixture estimator's bias depended on $\muc1$---when $\muc1=0$, the bias was slight, and when $\muc1=0.3$, the bias was more substantial. This result is somewhat surprising, because when $\muc1=0$, there was no separation between the principal strata in the control group, and previous literature \citep{griffin2008application} has identified this scenario as a particular challenge for mixture modeling.
% The \textsc{psw} estimator was also biased, due, presumably, to the violation of Principal Ignorability, Assumption \ref{ass:PI}, inherent in the data generating model. On the other hand, the \textsc{psw} estimator had easily the lowest sampling variance of the three estimators.

% The middle panel of Figure \ref{fig:boxplots} and the fifth line of Table \ref{tab:coverage} show results for when the residuals in \eqref{eq:y-sim} are uniformly distributed.
% In this case, \textsc{geepers} and \textsc{psw} estimators, which do not assume normality, behaved in a similar fashion as in the normal case, but the mixture estimator showed a bimodal pattern--sometimes over-estimating $\eff1$ and sometimes underestimating, but rarely estimating $\eff1$ accurately.
% Credible intervals from the mixture model exhibited severe undercoverage, while \textsc{geepers} confidence intervals achieved at least their nominal level.

% %\sloppy
% The rightmost panel of Figure \ref{fig:boxplots} returns to the case of normal residuals---and only shows results when $\muc1=0.3$ and $\alpha=0.5$---but introduces interactions in the data generating model between covariates and either $\st$, $Z$, or both, violating Assumption \ref{ass:rci}.
% %Presence of interactions could undermine the \textsc{geepers} by inducing a relationship between principal scores and potential outcomes, even after adjusting the outcomes for covariate effects.
% Fortunately, these interactions made little difference when $\alpha=0.5$. However, Table \ref{tab:coverage} shows that when $\alpha=0$ or 0.2, and interactions between $\bx$ and $Z$ were present (i.e., $\gamma_3\ne 0$), \textsc{geepers} 95\% confidence intervals tended to under-cover. Nevertheless, coverage of \textsc{geepers} intervals was nearly always substantially higher than coverage of corresponding 95\% credible intervals from mixture models.%\footnote{}
% >>>>>>> overleaf-2025-06-12-1948

@@

\subsubsection{Principal Stratification and Alternatives}
The goal of principal strafication in this context is to determine if assignment to the ASSISTments condition affected students who tend to request frequent bottom-out hints differently than students who tended to request fewer bottom-out hints.
Is this the right question to ask?

In a similar context, \citet{sales2021student} suggests two alternative appproaches: the average effect of requesting bottom-out hints on posttest scores for students in the ASSISTments condition, and estimating natural direct and indirect effects of assignment to the ASSISTments condition, treating bottom-out hint requests as a mediator.
The former approach is an observational study: although students were randomized to the ASSISTments condition, bottom-out hint requests were not randomized. Therefore, estimating the effect of bottom-out hints requires careful control of confounding.
Unfortunately, confounding is a major concern in this case, since, presumably, students only requested bottom-out hints if they were struggling with the material, and students who struggled with practice problems are likely to score lower on the posttest, on average, than those who needed less help during practice.
The best available measure of students' baseline ability, their score on the 10-item pretest, is likely insufficient.
That said, a plausible estimate of the average effect of bottom-out hint requests would be a valuable contribution to the literature, and a suitably innovative observational design \citep[e.g.][]{beck} may bear fruit.

Another arguable disadvantage of the observational study approach is that it ignores the larger context of the experiment, asking whether bottom-out hints affect posttest scores, rather than asking what role they play in ASSISTments' overall effectiveness.
A second approach, mediation analysis, contextualizes the bottom-out effect within the larger experiment.
It, too, relies on the same non-confounding assumption as an observational study, and also suffers from an additional challenge: mediation analysis relies on a ``common support'' condition that ``the mediator state must not be a deterministic function of the treatment'' \citep[p. 225]{celli2022causal}.
However, bottom-out hints are unavailable in the other three arms of the study, violating this condition (\citealt{sales2021student} offers a solution, which we find unconvincing).

A principal stratification approach treats bottom-out hint requests as a type of compliance.
In other words, if we consider full implementation of the ASSISTments condition to include frequent bottom-out hint requests, students assigned to ASSISTments but requests few bottom-out hints is only partially complying with their treatment assignment.
Then the question is if students who do not fully comply with their treatment assignment still experiance an effect.
Conversely, might students who only partially comply experience a larger positive effect?
That would be the case if either bottom-out hints were harmful, or if the treatment was better targeted toward the type of student who tends to need them less.

Principal score weighting depends on the assumption of principal ignorability, which is equivalent to the no-confounding assumption of the observational study and mediation analysis.
The mixture model and \textsc{geepers} estimators do not rely on that assumption, instead relying on parametric models and residualized covariate independence, respectively.
Finally, the instrumental variables estimator of \citet{air} relies on the ``exclusion restriction'' that, for students who request few bottom-out hints, there is no effect of randomization to the ASSISTments condition.
Since bottom-out hints are only one of several features of ASSISTments, this assumption is untenable.

\subsubsection{Data on Bottom-Out Hints and Covariates}
